---
title: "Getting started with CellOnTools"
author: "Di Wang"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting started with CellOnTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Optional dependencies — chunks that need them are evaluated only when present.
has_oi  <- requireNamespace("ontologyIndex", quietly = TRUE)
has_os  <- requireNamespace("ontologySimilarity", quietly = TRUE)
has_cp  <- requireNamespace("clusterProfiler", quietly = TRUE)
```

## Introduction

The [Cell Ontology](https://obofoundry.org/ontology/cl.html) (CL) is a
structured, species-neutral controlled vocabulary of cell types, organised as a
directed acyclic graph (DAG) in which each term is linked to more general
parent terms. **CellOnTools** provides a consistent, well-validated R interface
to common CL operations together with marker-gene over-representation analysis:

* load the fixed Cell Ontology release or an explicitly selected legacy/local
  release;
* convert between CL identifiers and labels, and search labels;
* query ancestors, descendants, depth, and common ancestors;
* extract and plot hierarchy subgraphs;
* compute Resnik / Lin semantic similarity;
* roll fine-grained terms up to coarser shared ancestors;
* map free-text cell-type names to CL terms via the Ontology Lookup Service;
* test gene sets for enrichment of cell-type markers using bundled healthy
  human and mouse CellMarkerAccordion annotations.

```{r library}
library(CellOnTools)
```

## Choosing a workflow

CellOnTools functions are designed to be combined rather than used in isolation.
The table below is a useful starting point:

| If you have... | Use... | Typical output |
|---|---|---|
| Reviewed CL IDs | `CLid2label()`, `CLdepth()`, `CLancestors()` | Human-readable labels, specificity summaries, parent terms |
| Reviewed cell-type labels | `CLlabel2id()` | Stable CL IDs for metadata tables |
| Shorthand or free-text labels | `CLsearchLabel()`, `CLmap()`, `CLmapInteractive()` | Search evidence or OLS-ranked candidate CL terms |
| Many fine-grained annotations | `CLsuggestResolution()`, `CLrollup()` | A coarser, ontology-aware annotation layer |
| Several annotated clusters | `CLsimilarityMatrix()`, `CLcommonAncestor()` | Pairwise similarity and shared biological neighbourhoods |
| Marker genes per cluster | `CLenricher()`, `CLcompareCluster()` | Marker-based evidence for candidate CL terms |

## Ancestor count versus graph distance

`CLdepth()` reports the **CL-only ancestor count**: the number of distinct
proper `CL:*` ancestors of a term. Imported ontology nodes are excluded. Because
CL is a DAG, this specificity measure is not the number of graph edges to a
root.

Neighbourhood queries use a separate and literal distance measure. The
`max_hops` argument in `CLancestors()`, `CLdescendants()`, `CLhierarchy()`, and
`CLhierarchyPlot()` limits the number of direct CL parent/child edges. Roll-up
uses `max_candidate_ancestor_count` as a specificity cap, not as distance.
CL-only filtering does not automatically remove obsolete CL terms; use the
loaded object's `obsolete` field when active-term-only analysis is required.

## Loading the Cell Ontology

In normal use you load the ontology once and reuse the object. `CLload()` uses
the fixed, checksum-verified `2026-06-08` release by default and caches it in
the user cache directory:

```{r load-network, eval = FALSE, purl = FALSE}
# Fixed 2026-06-08 release (default):
clData <- CLload()

# Or manage the same versioned file explicitly:
CLdownload(dest_file = "cl.obo")
clData <- CLload(local_obo = "cl.obo", prefer_local = TRUE)

# Legacy AnnotationHub selection remains explicit:
clData_2023 <- CLload(yearAdded = "2023")
```

Both the default cache and `CLdownload()` verify the release header and official
MD5 before parsing. The returned object records `ontology_release` and
`ontology_source` attributes for audit trails.

To keep this vignette fast and free of network access, the examples below use a
small, hand-built ontology that mimics the structure of CL:

```{r toy-ontology}
ids <- c("CL:0000000", "CL:0000100", "CL:0000200", "CL:0000300",
         "CL:0000400", "CL:0000350", "CL:0000700")
name <- stats::setNames(
  c("cell", "immune cell", "lymphocyte", "T cell",
    "B cell", "natural killer cell", "NKT cell"),
  ids
)
parents <- list(
  "CL:0000000" = character(0),
  "CL:0000100" = "CL:0000000",
  "CL:0000200" = "CL:0000100",
  "CL:0000300" = "CL:0000200",
  "CL:0000400" = "CL:0000200",
  "CL:0000350" = "CL:0000200",
  "CL:0000700" = c("CL:0000300", "CL:0000350")  # NKT cell: two parents
)
children <- list(
  "CL:0000000" = "CL:0000100",
  "CL:0000100" = "CL:0000200",
  "CL:0000200" = c("CL:0000300", "CL:0000400", "CL:0000350"),
  "CL:0000300" = "CL:0000700",
  "CL:0000400" = character(0),
  "CL:0000350" = "CL:0000700",
  "CL:0000700" = character(0)
)
ancestors <- list(
  "CL:0000000" = "CL:0000000",
  "CL:0000100" = c("CL:0000100", "CL:0000000"),
  "CL:0000200" = c("CL:0000200", "CL:0000100", "CL:0000000"),
  "CL:0000300" = c("CL:0000300", "CL:0000200", "CL:0000100", "CL:0000000"),
  "CL:0000400" = c("CL:0000400", "CL:0000200", "CL:0000100", "CL:0000000"),
  "CL:0000350" = c("CL:0000350", "CL:0000200", "CL:0000100", "CL:0000000"),
  "CL:0000700" = c("CL:0000700", "CL:0000300", "CL:0000350",
                   "CL:0000200", "CL:0000100", "CL:0000000")
)
clData <- structure(
  list(id = ids, name = name, parents = parents,
       children = children, ancestors = ancestors),
  class = "ontology_index"
)
```

## Converting and searching labels

These functions operate directly on the loaded object and require no optional
packages:

```{r convert}
CLid2label(c("CL:0000300", "CL:0000400"), clData)
CLlabel2id("T cell", clData)
CLsearchLabel("cell", clData)
```

Invalid-format and unknown identifiers return `NA` with a single aggregated
warning per problem class (use `strict = TRUE` to error instead):

```{r convert-na}
suppressWarnings(CLid2label(c("CL:0000300", "CL:9999999", "not_an_id"), clData))
```

In a real annotation table, use `strict = TRUE` when reviewed labels are
expected to exist in the ontology. This fails early if a label has drifted, was
misspelled, or belongs to a different ontology release:

```{r strict-real, eval = FALSE, purl = FALSE}
annotation_dictionary <- data.frame(
  cluster_label = c("B", "NK", "Platelet"),
  reviewed_cl_label = c("B cell", "natural killer cell", "platelet")
)

annotation_dictionary$cl_id <- CLlabel2id(
  annotation_dictionary$reviewed_cl_label,
  clData,
  strict = TRUE
)
```

## Ancestors, descendants and depth

```{r ancestry, purl = FALSE}
if (has_oi) {
  CLdepth(c("CL:0000300", "CL:0000700"), clData)
  CLancestors("CL:0000700", clData)
  CLdescendants("CL:0000200", clData, max_hops = 1)
}
```

The NKT cell (`CL:0000700`) has two parents. Its ancestor count (5) reflects the
size of its CL ancestor set, while `max_hops = 1` always means direct edges.

## Common ancestors and roll-up

`CLcommonAncestor()` returns shared ancestors; with `most_specific = TRUE` it
returns the most granular shared node(s):

```{r common, purl = FALSE}
if (has_oi) {
  CLcommonAncestor(c("CL:0000300", "CL:0000400"), clData,
                   most_specific = TRUE)
}
```

`CLrollup()` collapses a set of terms onto their most specific shared ancestors,
subject to a minimum group size — useful for harmonising fine-grained cluster
annotations to a coarser, comparable resolution:

```{r rollup, purl = FALSE}
if (has_oi) {
  rolled <- CLrollup(
    c("CL:0000300", "CL:0000400", "CL:0000700"),
    clData, min_group_size = 2, verbose = FALSE
  )
  rolled$mapping[, c("original_label", "rolled_label", "was_rolled")]
}
```

`CLsuggestResolution()` sweeps the roll-up parameters and recommends the setting
whose number of groups is closest to a target.

For single-cell annotations, this is useful when a fine-grained annotation layer
is too detailed for a figure or cross-dataset comparison. For example, CD4 T,
CD8 T, B, and NK annotations can be kept as reviewed CL IDs while a second
metadata column stores their broader roll-up term, such as `lymphocyte`.

## Semantic similarity

Resnik and Lin similarities are available when **ontologySimilarity** is
installed. Pre-computing the information content makes repeated comparisons
cheaper:

```{r similarity, purl = FALSE}
if (has_os) {
  ic <- ontologySimilarity::descendants_IC(clData)
  CLsimilarity("CL:0000300", "CL:0000400", clData,
               method = "lin", information_content = ic)
  CLsimilarityMatrix(c("CL:0000300", "CL:0000400", "CL:0000350"),
                     clData = clData, information_content = ic, verbose = FALSE)
}
```

`CLsimilarity()` and `CLsimilarityMatrix()` are helpful for asking whether two
annotations are close in CL even when their labels are not identical. Lin
similarity is bounded between 0 and 1, which makes it convenient for summaries
and heatmaps; Resnik similarity is often useful when you want an
information-content score for the shared ancestor itself.

## Hierarchy plots

`CLhierarchy()` returns graph-ready `nodes`/`edges` data frames, and
`CLhierarchyPlot()` renders them with **ggraph** (highlighting query terms).
Both require the relevant graph packages:

```{r plot, eval = FALSE, purl = FALSE}
CLhierarchyPlot(c("CL:0000300", "CL:0000400"), clData)
```

## Marker genes and enrichment

The package bundles healthy human and mouse positive-marker tables linking cell
types to active terms in CL release `2026-06-08`. Obsolete source IDs are
migrated through their official `replaced_by` values. Source database, tissue,
and weighting fields are not retained, so enrichment is pooled and unweighted:

```{r markers}
human <- CLmarkers("human")
head(human)
```

Given a gene set, `CLenricher()` tests for over-representation of cell-type
markers via **clusterProfiler**:

```{r enrich, purl = FALSE}
if (has_cp) {
  res <- CLenricher(
    c("CD3D", "CD3E", "CD8A", "CD8B", "CD4", "IL7R", "CCR7"),
    geneType = "symbol", species = "human"
  )
  if (!is.null(res)) {
    head(as.data.frame(res)[, c("ID", "Description", "p.adjust", "Count")], 5)
  }
}
```

The enrichment universe is part of the analysis definition. The default
`universe = NULL` uses all genes present in the bundled marker database for the
selected species. For a study-specific assay, pass the genes that could have
been detected—for example, `rownames(seurat_object)` for the unfiltered feature
space—so that the enrichment background matches the experiment. Record this
choice because changing the universe can alter both the number and ranking of
enriched terms.

`CLcompareCluster()` extends enrichment to several gene sets at once. This is
the usual shape of a single-cell workflow after marker detection:

```{r comparecluster-small, purl = FALSE}
if (has_cp) {
  cluster_markers <- list(
    "T-like"    = c("CD3D", "CD3E", "CD8A", "IL7R"),
    "B-like"    = c("MS4A1", "CD79A", "CD79B", "CD74"),
    "Mono-like" = c("LYZ", "S100A8", "S100A9", "FCN1")
  )

  cc <- CLcompareCluster(cluster_markers, geneType = "symbol", species = "human")
  cc_df <- as.data.frame(cc)
  if (nrow(cc_df)) {
    head(cc_df[, c("Cluster", "ID", "Description", "p.adjust")], 8)
  }
}
```

When no term passes the requested cutoffs, `CLenricher()` may return `NULL` and
`CLcompareCluster()` may contain no rows. In scripted workflows, check this
before plotting.

## Case-study lessons from pbmc3k

The package documentation was expanded using a manuscript-style pbmc3k analysis
that connected published Seurat annotations to Cell Ontology, explored ontology
relationships, rolled annotations to a coarser layer, and compared the reviewed
labels with marker-gene enrichment evidence. The 2,700-cell raw object is used
only for QC context; annotation analyses use the published 2,638-cell
`pbmc3k.final` object. The example is therefore a standardisation and
concordance case study, not an end-to-end clustering or independent automated
annotation benchmark.

Key results from that run and the current background-sensitivity check:

| Quantity | Value |
|---|---:|
| Raw pbmc3k dimensions | 13,714 genes × 2,700 cells |
| Published `pbmc3k.final` dimensions | 13,714 genes × 2,638 cells |
| Seurat clusters / reviewed labels | 9 / 9 |
| Reviewed labels resolved to CL | 9 / 9 |
| Locked roll-up setting | Preserve platelet; `min_group_size = 2`, ancestor-count cap 8 for the other 8 terms |
| Reviewed annotations after roll-up | 9 to 4 final groups (3 leukocyte groups plus platelet) |
| Seurat marker rows | 3,456 |
| `CLcompareCluster()` rows, default marker universe | 674 |
| `CLcompareCluster()` rows, separate 13,714-gene universe sensitivity analysis | 663 |
| Exact CL-ID matches, depth-prioritised exploratory rule | 1 / 9 |
| Median Lin similarity, depth-prioritised exploratory rule | 0.875 |
| Ancestor-descendant-comparable selections | 9 / 9 |

The exact-match number is intentionally not the only success criterion. Marker
enrichment often returns a nearby parent, child, or sibling term. The current
script uses an exploratory rule that retains candidates within ten-fold of the
smallest adjusted P value and then prefers the greatest CL-only ancestor count.
For example, the reviewed annotation `CD14-low, CD16-positive monocyte` was
supported by `non-classical monocyte` with high Lin similarity. The quoted
674-row and 0.875 results use the package's default marker-database universe;
the separate 663-row sensitivity result shows why the background must be
reported.

The recommended pattern is:

1. keep a versioned, reviewed annotation dictionary with explicit CL IDs and
   the search evidence used to choose them;
2. store the original labels, CL IDs, and CL labels together in cell metadata;
3. use marker enrichment as evidence and quality control, with an explicitly
   declared universe;
4. use semantic similarity to quantify near-matches rather than relying only on
   exact ID equality;
5. when evaluating marker-based label recovery, derive markers from blinded
   cluster IDs and join the reference labels only after predictions are locked.

## Mapping free-text names to CL terms

`CLmap()` (batch) and `CLmapInteractive()` (console review) resolve free-text
cell-type names to CL terms through the EBI Ontology Lookup Service. These
require network access and the **rols** package. Mapping proceeds only when OLS
reports the same `2026-06-08` release; otherwise it stops before searching:

```{r map-network, eval = FALSE, purl = FALSE}
CLmap(c("T cell", "B cell", "classical monocyte"), returnType = "id")
```

Because OLS is network-dependent, manuscript and pipeline code should record
failures rather than allowing one API outage to erase local results:

```{r map-guard, eval = FALSE, purl = FALSE}
ols_candidates <- tryCatch(
  CLmap(c("naive CD4-positive alpha-beta T cell", "CD16-positive monocyte"),
        max_results = 3, returnType = "all", verbose = FALSE),
  error = function(e) {
    data.frame(error_message = conditionMessage(e))
  }
)
```

## Session information

```{r sessioninfo}
sessionInfo()
```
