---
title: "Getting started with CellOnTools"
author: "Di Wang"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting started with CellOnTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Optional dependencies — chunks that need them are evaluated only when present.
has_oi  <- requireNamespace("ontologyIndex", quietly = TRUE)
has_os  <- requireNamespace("ontologySimilarity", quietly = TRUE)
has_cp  <- requireNamespace("clusterProfiler", quietly = TRUE)
```

## Introduction

The [Cell Ontology](https://obofoundry.org/ontology/cl.html) (CL) is a
structured, species-neutral controlled vocabulary of cell types, organised as a
directed acyclic graph (DAG) in which each term is linked to more general
parent terms. **CellOnTools** provides a consistent, well-validated R interface
to common CL operations together with marker-gene over-representation analysis:

* load the ontology from Bioconductor's AnnotationHub or a local OBO file;
* convert between CL identifiers and labels, and search labels;
* query ancestors, descendants, depth, and common ancestors;
* extract and plot hierarchy subgraphs;
* compute Resnik / Lin semantic similarity;
* roll fine-grained terms up to coarser shared ancestors;
* map free-text cell-type names to CL terms via the Ontology Lookup Service;
* test gene sets for enrichment of cell-type markers using bundled healthy
  human and mouse CellMarkerAccordion annotations.

```{r library}
library(CellOnTools)
```

## A note on "depth"

Throughout the package, **depth is defined as the ancestor count**: the number
of distinct proper ancestors of a term (i.e. excluding the term itself). The
root has depth 0, its direct children have depth 1, and more specific terms have
larger depths. Because CL is a DAG, this is *not* the number of edges to the
root; it is the cardinality of the (transitive) ancestor set. All depth-aware
arguments (`max_ancestor_count`, etc.) use this convention.

## Loading the Cell Ontology

In normal use you load the ontology once and reuse the object. `CLload()` first
tries AnnotationHub and falls back to a local OBO file:

```{r load-network, eval = FALSE}
# From AnnotationHub (requires the AnnotationHub + S4Vectors packages):
clData <- CLload(yearAdded = "2023")

# Or download the latest release and load it locally:
CLdownload(dest_file = "cl.obo")
clData <- CLload(local_obo = "cl.obo", prefer_local = TRUE)
```

To keep this vignette fast and free of network access, the examples below use a
small, hand-built ontology that mimics the structure of CL:

```{r toy-ontology}
ids <- c("CL:0000000", "CL:0000100", "CL:0000200", "CL:0000300",
         "CL:0000400", "CL:0000350", "CL:0000700")
name <- stats::setNames(
  c("cell", "immune cell", "lymphocyte", "T cell",
    "B cell", "natural killer cell", "NKT cell"),
  ids
)
parents <- list(
  "CL:0000000" = character(0),
  "CL:0000100" = "CL:0000000",
  "CL:0000200" = "CL:0000100",
  "CL:0000300" = "CL:0000200",
  "CL:0000400" = "CL:0000200",
  "CL:0000350" = "CL:0000200",
  "CL:0000700" = c("CL:0000300", "CL:0000350")  # NKT cell: two parents
)
children <- list(
  "CL:0000000" = "CL:0000100",
  "CL:0000100" = "CL:0000200",
  "CL:0000200" = c("CL:0000300", "CL:0000400", "CL:0000350"),
  "CL:0000300" = "CL:0000700",
  "CL:0000400" = character(0),
  "CL:0000350" = "CL:0000700",
  "CL:0000700" = character(0)
)
ancestors <- list(
  "CL:0000000" = "CL:0000000",
  "CL:0000100" = c("CL:0000100", "CL:0000000"),
  "CL:0000200" = c("CL:0000200", "CL:0000100", "CL:0000000"),
  "CL:0000300" = c("CL:0000300", "CL:0000200", "CL:0000100", "CL:0000000"),
  "CL:0000400" = c("CL:0000400", "CL:0000200", "CL:0000100", "CL:0000000"),
  "CL:0000350" = c("CL:0000350", "CL:0000200", "CL:0000100", "CL:0000000"),
  "CL:0000700" = c("CL:0000700", "CL:0000300", "CL:0000350",
                   "CL:0000200", "CL:0000100", "CL:0000000")
)
clData <- structure(
  list(id = ids, name = name, parents = parents,
       children = children, ancestors = ancestors),
  class = "ontology_index"
)
```

## Converting and searching labels

These functions operate directly on the loaded object and require no optional
packages:

```{r convert}
CLid2label(c("CL:0000300", "CL:0000400"), clData)
CLlabel2id("T cell", clData)
CLsearchLabel("cell", clData)
```

Invalid-format and unknown identifiers return `NA` with a single aggregated
warning per problem class (use `strict = TRUE` to error instead):

```{r convert-na}
suppressWarnings(CLid2label(c("CL:0000300", "CL:9999999", "not_an_id"), clData))
```

## Ancestors, descendants and depth

```{r ancestry, eval = has_oi}
CLdepth(c("CL:0000300", "CL:0000700"), clData)
CLancestors("CL:0000700", clData)
CLdescendants("CL:0000200", clData, max_descendant_ancestor_count = 1)
```

The NKT cell (`CL:0000700`) has two parents, so its depth (5) reflects the size
of its ancestor set, not a single path length.

## Common ancestors and roll-up

`CLcommonAncestor()` returns shared ancestors; with `most_specific = TRUE` it
returns the most granular shared node(s):

```{r common, eval = has_oi}
CLcommonAncestor(c("CL:0000300", "CL:0000400"), clData, most_specific = TRUE)
```

`CLrollup()` collapses a set of terms onto their most specific shared ancestors,
subject to a minimum group size — useful for harmonising fine-grained cluster
annotations to a coarser, comparable resolution:

```{r rollup, eval = has_oi}
rolled <- CLrollup(
  c("CL:0000300", "CL:0000400", "CL:0000700"),
  clData, min_group_size = 2, verbose = FALSE
)
rolled$mapping[, c("original_label", "rolled_label", "was_rolled")]
```

`CLsuggestResolution()` sweeps the roll-up parameters and recommends the setting
whose number of groups is closest to a target.

## Semantic similarity

Resnik and Lin similarities are available when **ontologySimilarity** is
installed. Pre-computing the information content makes repeated comparisons
cheaper:

```{r similarity, eval = has_os}
ic <- ontologySimilarity::descendants_IC(clData)
CLsimilarity("CL:0000300", "CL:0000400", clData,
             method = "lin", information_content = ic)
CLsimilarityMatrix(c("CL:0000300", "CL:0000400", "CL:0000350"),
                   clData = clData, information_content = ic, verbose = FALSE)
```

## Hierarchy plots

`CLhierarchy()` returns graph-ready `nodes`/`edges` data frames, and
`CLhierarchyPlot()` renders them with **ggraph** (highlighting query terms).
Both require the relevant graph packages:

```{r plot, eval = FALSE}
CLhierarchyPlot(c("CL:0000300", "CL:0000400"), clData)
```

## Marker genes and enrichment

The package bundles healthy human and mouse marker tables linking cell types to
CL identifiers and marker genes:

```{r markers}
human <- CLmarkers("human")
head(human)
```

Given a gene set, `CLenricher()` tests for over-representation of cell-type
markers via **clusterProfiler**:

```{r enrich, eval = has_cp}
res <- CLenricher(
  c("CD3D", "CD3E", "CD8A", "CD8B", "CD4", "IL7R", "CCR7"),
  geneType = "symbol", species = "human"
)
head(as.data.frame(res)[, c("ID", "Description", "p.adjust", "Count")], 5)
```

`CLcompareCluster()` extends this to several gene sets at once (e.g. marker
genes per cluster from a single-cell experiment), returning a
`compareClusterResult` that integrates with **enrichplot** visualisations.

## Mapping free-text names to CL terms

`CLmap()` (batch) and `CLmapInteractive()` (console review) resolve free-text
cell-type names to CL terms through the EBI Ontology Lookup Service. These
require network access and the **rols** package:

```{r map-network, eval = FALSE}
CLmap(c("T cell", "B cell", "classical monocyte"), returnType = "id")
```

## Session information

```{r sessioninfo}
sessionInfo()
```
