---
title: "Getting started with CellOnTools"
author: "Di Wang"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting started with CellOnTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Optional dependencies — chunks that need them are evaluated only when present.
has_oi  <- requireNamespace("ontologyIndex", quietly = TRUE)
has_os  <- requireNamespace("ontologySimilarity", quietly = TRUE)
has_cp  <- requireNamespace("clusterProfiler", quietly = TRUE)
```

## Introduction

The [Cell Ontology](https://obofoundry.org/ontology/cl.html) (CL) is a
structured, species-neutral controlled vocabulary of cell types, organised as a
directed acyclic graph (DAG) in which each term is linked to more general
parent terms. **CellOnTools** provides a consistent, well-validated R interface
to common CL operations together with marker-gene over-representation analysis:

* load the ontology from Bioconductor's AnnotationHub or a local OBO file;
* convert between CL identifiers and labels, and search labels;
* query ancestors, descendants, depth, and common ancestors;
* extract and plot hierarchy subgraphs;
* compute Resnik / Lin semantic similarity;
* roll fine-grained terms up to coarser shared ancestors;
* map free-text cell-type names to CL terms via the Ontology Lookup Service;
* test gene sets for enrichment of cell-type markers using bundled healthy
  human and mouse CellMarkerAccordion annotations.

```{r library}
library(CellOnTools)
```

## Choosing a workflow

CellOnTools functions are designed to be combined rather than used in isolation.
The table below is a useful starting point:

| If you have... | Use... | Typical output |
|---|---|---|
| Reviewed CL IDs | `CLid2label()`, `CLdepth()`, `CLancestors()` | Human-readable labels, specificity summaries, parent terms |
| Reviewed cell-type labels | `CLlabel2id()` | Stable CL IDs for metadata tables |
| Shorthand or free-text labels | `CLsearchLabel()`, `CLmap()`, `CLmapInteractive()` | Search evidence or OLS-ranked candidate CL terms |
| Many fine-grained annotations | `CLsuggestResolution()`, `CLrollup()` | A coarser, ontology-aware annotation layer |
| Several annotated clusters | `CLsimilarityMatrix()`, `CLcommonAncestor()` | Pairwise similarity and shared biological neighbourhoods |
| Marker genes per cluster | `CLenricher()`, `CLcompareCluster()` | Marker-based evidence for candidate CL terms |

## A note on "depth"

Throughout the package, **depth is defined as the ancestor count**: the number
of distinct proper ancestors of a term (i.e. excluding the term itself). The
root has depth 0, its direct children have depth 1, and more specific terms have
larger depths. Because CL is a DAG, this is *not* the number of edges to the
root; it is the cardinality of the (transitive) ancestor set. All depth-aware
arguments (`max_ancestor_count`, etc.) use this convention.

## Loading the Cell Ontology

In normal use you load the ontology once and reuse the object. `CLload()` first
tries AnnotationHub and falls back to a local OBO file:

```{r load-network, eval = FALSE}
# From AnnotationHub (requires the AnnotationHub + S4Vectors packages):
clData <- CLload(yearAdded = "2023")

# Or download the latest release and load it locally:
CLdownload(dest_file = "cl.obo")
clData <- CLload(local_obo = "cl.obo", prefer_local = TRUE)
```

For reproducible analyses, the local OBO path is often preferable: it records
exactly which ontology file was used, avoids AnnotationHub availability changes,
and makes reruns easier to audit.

To keep this vignette fast and free of network access, the examples below use a
small, hand-built ontology that mimics the structure of CL:

```{r toy-ontology}
ids <- c("CL:0000000", "CL:0000100", "CL:0000200", "CL:0000300",
         "CL:0000400", "CL:0000350", "CL:0000700")
name <- stats::setNames(
  c("cell", "immune cell", "lymphocyte", "T cell",
    "B cell", "natural killer cell", "NKT cell"),
  ids
)
parents <- list(
  "CL:0000000" = character(0),
  "CL:0000100" = "CL:0000000",
  "CL:0000200" = "CL:0000100",
  "CL:0000300" = "CL:0000200",
  "CL:0000400" = "CL:0000200",
  "CL:0000350" = "CL:0000200",
  "CL:0000700" = c("CL:0000300", "CL:0000350")  # NKT cell: two parents
)
children <- list(
  "CL:0000000" = "CL:0000100",
  "CL:0000100" = "CL:0000200",
  "CL:0000200" = c("CL:0000300", "CL:0000400", "CL:0000350"),
  "CL:0000300" = "CL:0000700",
  "CL:0000400" = character(0),
  "CL:0000350" = "CL:0000700",
  "CL:0000700" = character(0)
)
ancestors <- list(
  "CL:0000000" = "CL:0000000",
  "CL:0000100" = c("CL:0000100", "CL:0000000"),
  "CL:0000200" = c("CL:0000200", "CL:0000100", "CL:0000000"),
  "CL:0000300" = c("CL:0000300", "CL:0000200", "CL:0000100", "CL:0000000"),
  "CL:0000400" = c("CL:0000400", "CL:0000200", "CL:0000100", "CL:0000000"),
  "CL:0000350" = c("CL:0000350", "CL:0000200", "CL:0000100", "CL:0000000"),
  "CL:0000700" = c("CL:0000700", "CL:0000300", "CL:0000350",
                   "CL:0000200", "CL:0000100", "CL:0000000")
)
clData <- structure(
  list(id = ids, name = name, parents = parents,
       children = children, ancestors = ancestors),
  class = "ontology_index"
)
```

## Converting and searching labels

These functions operate directly on the loaded object and require no optional
packages:

```{r convert}
CLid2label(c("CL:0000300", "CL:0000400"), clData)
CLlabel2id("T cell", clData)
CLsearchLabel("cell", clData)
```

Invalid-format and unknown identifiers return `NA` with a single aggregated
warning per problem class (use `strict = TRUE` to error instead):

```{r convert-na}
suppressWarnings(CLid2label(c("CL:0000300", "CL:9999999", "not_an_id"), clData))
```

In a real annotation table, use `strict = TRUE` when reviewed labels are
expected to exist in the ontology. This fails early if a label has drifted, was
misspelled, or belongs to a different ontology release:

```{r strict-real, eval = FALSE}
annotation_dictionary <- data.frame(
  cluster_label = c("B", "NK", "Platelet"),
  reviewed_cl_label = c("B cell", "natural killer cell", "platelet")
)

annotation_dictionary$cl_id <- CLlabel2id(
  annotation_dictionary$reviewed_cl_label,
  clData,
  strict = TRUE
)
```

## Ancestors, descendants and depth

```{r ancestry, eval = has_oi}
CLdepth(c("CL:0000300", "CL:0000700"), clData)
CLancestors("CL:0000700", clData)
CLdescendants("CL:0000200", clData, max_descendant_ancestor_count = 1)
```

The NKT cell (`CL:0000700`) has two parents, so its depth (5) reflects the size
of its ancestor set, not a single path length.

## Common ancestors and roll-up

`CLcommonAncestor()` returns shared ancestors; with `most_specific = TRUE` it
returns the most granular shared node(s):

```{r common, eval = has_oi}
CLcommonAncestor(c("CL:0000300", "CL:0000400"), clData, most_specific = TRUE)
```

`CLrollup()` collapses a set of terms onto their most specific shared ancestors,
subject to a minimum group size — useful for harmonising fine-grained cluster
annotations to a coarser, comparable resolution:

```{r rollup, eval = has_oi}
rolled <- CLrollup(
  c("CL:0000300", "CL:0000400", "CL:0000700"),
  clData, min_group_size = 2, verbose = FALSE
)
rolled$mapping[, c("original_label", "rolled_label", "was_rolled")]
```

`CLsuggestResolution()` sweeps the roll-up parameters and recommends the setting
whose number of groups is closest to a target.

For single-cell annotations, this is useful when a fine-grained annotation layer
is too detailed for a figure or cross-dataset comparison. For example, CD4 T,
CD8 T, B, and NK annotations can be kept as reviewed CL IDs while a second
metadata column stores their broader roll-up term, such as `lymphocyte`.

## Semantic similarity

Resnik and Lin similarities are available when **ontologySimilarity** is
installed. Pre-computing the information content makes repeated comparisons
cheaper:

```{r similarity, eval = has_os}
ic <- ontologySimilarity::descendants_IC(clData)
CLsimilarity("CL:0000300", "CL:0000400", clData,
             method = "lin", information_content = ic)
CLsimilarityMatrix(c("CL:0000300", "CL:0000400", "CL:0000350"),
                   clData = clData, information_content = ic, verbose = FALSE)
```

`CLsimilarity()` and `CLsimilarityMatrix()` are helpful for asking whether two
annotations are close in CL even when their labels are not identical. Lin
similarity is bounded between 0 and 1, which makes it convenient for summaries
and heatmaps; Resnik similarity is often useful when you want an
information-content score for the shared ancestor itself.

## Hierarchy plots

`CLhierarchy()` returns graph-ready `nodes`/`edges` data frames, and
`CLhierarchyPlot()` renders them with **ggraph** (highlighting query terms).
Both require the relevant graph packages:

```{r plot, eval = FALSE}
CLhierarchyPlot(c("CL:0000300", "CL:0000400"), clData)
```

## Marker genes and enrichment

The package bundles healthy human and mouse marker tables linking cell types to
CL identifiers and marker genes:

```{r markers}
human <- CLmarkers("human")
head(human)
```

Given a gene set, `CLenricher()` tests for over-representation of cell-type
markers via **clusterProfiler**:

```{r enrich, eval = has_cp}
res <- CLenricher(
  c("CD3D", "CD3E", "CD8A", "CD8B", "CD4", "IL7R", "CCR7"),
  geneType = "symbol", species = "human"
)
head(as.data.frame(res)[, c("ID", "Description", "p.adjust", "Count")], 5)
```

The default `universe = NULL` uses all genes present in the bundled marker
database for the selected species. If your experiment measured a restricted
panel, pass the measured genes as `universe` so the enrichment background
matches the assay.

`CLcompareCluster()` extends enrichment to several gene sets at once. This is
the usual shape of a single-cell workflow after marker detection:

```{r comparecluster-small, eval = has_cp}
cluster_markers <- list(
  "T-like"    = c("CD3D", "CD3E", "CD8A", "IL7R"),
  "B-like"    = c("MS4A1", "CD79A", "CD79B", "CD74"),
  "Mono-like" = c("LYZ", "S100A8", "S100A9", "FCN1")
)

cc <- CLcompareCluster(cluster_markers, geneType = "symbol", species = "human")
head(as.data.frame(cc)[, c("Cluster", "ID", "Description", "p.adjust")], 8)
```

When no term passes the requested cutoffs, `CLenricher()` may return `NULL` and
`CLcompareCluster()` may contain no rows. In scripted workflows, check this
before plotting.

## Case-study lessons from pbmc3k

The package documentation was expanded using a manuscript-style pbmc3k analysis
that connected Seurat annotations to Cell Ontology, explored ontology
relationships, rolled annotations to a coarser layer, and compared the reviewed
labels with marker-gene enrichment evidence.

Key results from that run:

| Quantity | Value |
|---|---:|
| Processed pbmc3k cells | 2,638 |
| Seurat annotation labels resolved to CL | 9 / 9 |
| CL terms loaded from the local OBO file | 18,460 |
| Reviewed annotations after roll-up | 9 to 3 groups |
| Seurat marker rows used for enrichment | 3,456 |
| `CLcompareCluster()` enrichment rows | 674 |
| Top enrichment exact CL-ID matches | 1 / 9 |
| Median Lin similarity between top enriched and reviewed terms | 0.790 |

The exact-match number is intentionally not the only success criterion. Marker
enrichment often returns a nearby parent, child, or sibling term. For example,
the reviewed annotation `CD14-low, CD16-positive monocyte` was supported by a
top enriched term `non-classical monocyte` with high Lin similarity. This is why
the recommended pattern is:

1. keep a reviewed annotation dictionary with explicit CL IDs;
2. store CL IDs and labels back into cell metadata;
3. use marker enrichment as evidence and quality control;
4. use semantic similarity to quantify near-matches rather than relying only on
   exact ID equality.

For a fuller single-cell workflow template, see:

```{r pbmc-vignette-link, eval = FALSE}
vignette("pbmc3k-workflow", package = "CellOnTools")
```

## Mapping free-text names to CL terms

`CLmap()` (batch) and `CLmapInteractive()` (console review) resolve free-text
cell-type names to CL terms through the EBI Ontology Lookup Service. These
require network access and the **rols** package:

```{r map-network, eval = FALSE}
CLmap(c("T cell", "B cell", "classical monocyte"), returnType = "id")
```

Because OLS is network-dependent, manuscript and pipeline code should record
failures rather than allowing one API outage to erase local results:

```{r map-guard, eval = FALSE}
ols_candidates <- tryCatch(
  CLmap(c("naive CD4-positive alpha-beta T cell", "CD16-positive monocyte"),
        max_results = 3, returnType = "all", verbose = FALSE),
  error = function(e) {
    data.frame(error_message = conditionMessage(e))
  }
)
```

## Session information

```{r sessioninfo}
sessionInfo()
```
