---
title: "PBMC3K annotation workflow with CellOnTools"
author: "Di Wang"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{PBMC3K annotation workflow with CellOnTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Purpose

This vignette shows a manuscript-style workflow for using **CellOnTools** with a
single-cell object. It is based on a pbmc3k case study in which published Seurat
annotations were standardised to Cell Ontology (CL), written back to cell
metadata, rolled up to broader ontology terms, and compared with marker-gene
enrichment evidence.

The raw 2,700-cell object is used only to show the standard Seurat QC context;
all annotation analyses use the published 2,638-cell `pbmc3k.final` object.
Accordingly, this is a standardisation and concordance case study, not an
end-to-end benchmark of preprocessing, clustering, or independent automated
cell typing. A separate label-recovery evaluation should derive markers from
blinded cluster IDs and join reference labels only after predictions are
locked.

The full workflow chunks are intentionally not evaluated during package checks
because they require external pbmc3k objects, optional Bioconductor and plotting packages,
and network access for some mapping steps. Treat this vignette as a template;
the source repository also contains the standalone analysis script
`pbmc3k/pbmc3k_cellontools_case_study.R`.

The following small offline check is evaluated when the vignette is built, so
the document still verifies its CellOnTools installation without downloading
data or contacting an ontology service:

```{r offline-smoke, eval = TRUE, purl = TRUE}
human_markers <- CellOnTools::CLmarkers("human")
stopifnot(
  is.data.frame(human_markers),
  all(c("CL_ID", "marker_symbol") %in% names(human_markers))
)
```

```{r libraries, purl = FALSE}
library(CellOnTools)
library(Seurat)
library(ggplot2)
```

Recommended optional packages for the full workflow:

| Step | Packages |
|---|---|
| Single-cell object | `Seurat` |
| Local CL loading | `ontologyIndex` |
| Semantic similarity | `ontologySimilarity` |
| Enrichment | `clusterProfiler`, `enrichplot` |
| Hierarchy plotting | `ggraph`, `igraph`, `ggplot2` |
| OLS candidate search | `rols` |

## Load the ontology and pbmc3k

For a manuscript or pipeline, keep the package's fixed `2026-06-08` OBO file
locally so that the exact release is explicit and rerunnable:

```{r load, purl = FALSE}
out_dir <- "pbmc3k_results"
dir.create(out_dir, showWarnings = FALSE)

obo_file <- file.path(out_dir, "cl.obo")
if (!file.exists(obo_file)) {
  CLdownload(dest_file = obo_file, overwrite = TRUE)
}
cl_data <- CLload(local_obo = obo_file, prefer_local = TRUE, verbose = FALSE)
stopifnot(identical(attr(cl_data, "ontology_release"), "2026-06-08"))

# Load `pbmc_raw` and `pbmc` before this point. The standalone source-tree
# script uses SeuratData; the remainder of this vignette is data-source agnostic.
stopifnot(exists("pbmc_raw"), exists("pbmc"))
DefaultAssay(pbmc_raw) <- "RNA"
DefaultAssay(pbmc) <- "RNA"

pbmc_raw[["percent.mt"]] <- PercentageFeatureSet(pbmc_raw, pattern = "^MT-")
qc_keep <- pbmc_raw$nFeature_RNA > 200 &
  pbmc_raw$nFeature_RNA < 2500 &
  pbmc_raw$percent.mt < 5

stopifnot(
  nrow(pbmc_raw) == 13714L,
  ncol(pbmc_raw) == 2700L,
  sum(qc_keep) == 2638L,
  ncol(pbmc) == 2638L
)

# Reference-label grouping is used below for the concordance demonstration.
Idents(pbmc) <- pbmc$seurat_annotations
```

The raw and processed objects contain 13,714 genes. Exactly 2,638 of the 2,700
raw cells pass the tutorial thresholds shown above, matching the cell count in
`pbmc3k.final`. Loading an older SeuratData object under a newer Seurat release
may update its in-memory slots; record the Seurat, SeuratObject, SeuratData, and
CellOnTools versions with the analysis outputs.

## Build a reviewed CL dictionary

Start with a small table that keeps the original labels, search labels, and
reviewed CL labels separate. This makes the mapping auditable:

```{r reviewed-dictionary, purl = FALSE}
pbmc_terms <- data.frame(
  seurat_label = c("Naive CD4 T", "Memory CD4 T", "CD14+ Mono", "B", "CD8 T",
                   "FCGR3A+ Mono", "NK", "DC", "Platelet"),
  plot_label = c("Naive CD4 T", "Memory CD4 T", "CD14+ Mono", "B cell",
                 "CD8 T", "CD16+ Mono", "NK", "DC", "Platelet"),
  query_label = c("naive CD4-positive alpha-beta T cell",
                  "memory CD4-positive alpha-beta T cell",
                  "CD14-positive monocyte",
                  "B cell",
                  "CD8-positive alpha-beta T cell",
                  "CD16-positive monocyte",
                  "natural killer cell",
                  "dendritic cell",
                  "platelet"),
  reviewed_cl_label = c(
    "naive thymus-derived CD4-positive, alpha-beta T cell",
    "CD4-positive, alpha-beta memory T cell",
    "CD14-positive monocyte",
    "B cell",
    "CD8-positive, alpha-beta T cell",
    "CD14-low, CD16-positive monocyte",
    "natural killer cell",
    "dendritic cell",
    "platelet"
  ),
  standardisation_method = c(
    "CLmap", "CLmap", "CLmap", "CLmapInteractive", "CLmap",
    "CLmapInteractive", "CLmapInteractive", "CLmapInteractive",
    "CLsearchLabel"
  )
)
```

Use local search and OLS candidates to audit shorthand labels before finalising
the reviewed dictionary:

```{r search-and-map, purl = FALSE}
search_hits <- CLsearchLabel(
  c("CD4-positive", "monocyte", "natural killer cell",
    "dendritic cell", "platelet"),
  cl_data,
  max_results = 8
)

ols_candidates <- tryCatch(
  CLmap(pbmc_terms$query_label, max_results = 3,
        returnType = "all", verbose = FALSE),
  error = function(e) {
    data.frame(error_message = conditionMessage(e))
  }
)

utils::write.csv(search_hits,
                 file.path(out_dir, "pbmc3k_local_search_candidates.csv"),
                 row.names = FALSE)
utils::write.csv(ols_candidates,
                 file.path(out_dir, "pbmc3k_ols_candidates.csv"),
                 row.names = FALSE)
```

`CLmapInteractive()` can be used during console review to replace ambiguous
abbreviations with more informative searches—for example, `B` with `B cell` or
`FCGR3A+ Mono` with `CD16+ mono`. Those interactive decisions should not be
replayed automatically in a production run. Lock the accepted search term,
CL ID, method, reviewer, and ontology release in a version-controlled dictionary
and let the deterministic downstream analysis consume that dictionary.

Then resolve the reviewed labels to CL IDs and add ontology metadata:

```{r resolve-labels, purl = FALSE}
pbmc_terms$cl_id <- CLlabel2id(pbmc_terms$reviewed_cl_label,
                               cl_data, strict = TRUE)
pbmc_terms$cl_label <- CLid2label(pbmc_terms$cl_id, cl_data)
pbmc_terms$cl_depth <- CLdepth(pbmc_terms$cl_id, cl_data)
pbmc_terms$n_ancestors <- lengths(CLancestors(pbmc_terms$cl_id, cl_data))
pbmc_terms$n_descendants <- lengths(CLdescendants(pbmc_terms$cl_id, cl_data))
pbmc_terms$ontology_release <- attr(cl_data, "ontology_release")

stopifnot(!anyNA(pbmc_terms$cl_id), !anyDuplicated(pbmc_terms$seurat_label))
utils::write.csv(pbmc_terms,
                 file.path(out_dir, "pbmc3k_reviewed_cl_dictionary.csv"),
                 row.names = FALSE)
```

The reference run resolved all 9 labels:

| Seurat label | Cells | CL ID | CL label | Depth |
|---|---:|---|---|---:|
| Naive CD4 T | 697 | `CL:0000895` | naive thymus-derived CD4-positive, alpha-beta T cell | 18 |
| Memory CD4 T | 483 | `CL:0000897` | CD4-positive, alpha-beta memory T cell | 18 |
| CD14+ Mono | 480 | `CL:0001054` | CD14-positive monocyte | 16 |
| B | 344 | `CL:0000236` | B cell | 13 |
| CD8 T | 271 | `CL:0000625` | CD8-positive, alpha-beta T cell | 16 |
| FCGR3A+ Mono | 162 | `CL:0002396` | CD14-low, CD16-positive monocyte | 18 |
| NK | 155 | `CL:0000623` | natural killer cell | 14 |
| DC | 32 | `CL:0000451` | dendritic cell | 12 |
| Platelet | 14 | `CL:0000233` | platelet | 12 |

## Write CL terms back to cell metadata

Keep both the original annotation and the CL-standardised annotation. This
preserves provenance and makes downstream plotting straightforward:

```{r metadata, purl = FALSE}
id_by_label <- setNames(pbmc_terms$cl_id, pbmc_terms$seurat_label)
label_by_id <- setNames(pbmc_terms$cl_label, pbmc_terms$cl_id)
plot_by_id <- setNames(pbmc_terms$plot_label, pbmc_terms$cl_id)
depth_by_id <- setNames(pbmc_terms$cl_depth, pbmc_terms$cl_id)

pbmc$cellontools_cl_id <- unname(id_by_label[as.character(pbmc$seurat_annotations)])
pbmc$cellontools_cl_label <- unname(label_by_id[pbmc$cellontools_cl_id])
pbmc$cellontools_cl_plot_label <- unname(plot_by_id[pbmc$cellontools_cl_id])
pbmc$cellontools_cl_depth <- unname(depth_by_id[pbmc$cellontools_cl_id])

DimPlot(pbmc, group.by = "cellontools_cl_plot_label",
        label = TRUE, repel = TRUE) +
  NoLegend()
```

## Explore ontology relationships

Once annotations are CL IDs, you can inspect shared ancestors, hierarchy
subgraphs, and semantic similarity:

```{r relationships, purl = FALSE}
hier <- CLhierarchy(pbmc_terms$cl_id, cl_data,
                    include_ancestors = TRUE,
                    max_hops = NULL)

CLhierarchyPlot(pbmc_terms$cl_id, cl_data,
                include_ancestors = TRUE,
                max_hops = NULL,
                show_id = FALSE)

ic <- ontologySimilarity::descendants_IC(cl_data)
sim_lin <- CLsimilarityMatrix(pbmc_terms$cl_id, clData = cl_data,
                              method = "lin",
                              information_content = ic,
                              verbose = FALSE)

# Common ancestors may contain more than one maximally specific term in a DAG.
monocyte_common <- CLcommonAncestor(
  c("CL:0001054", "CL:0002396"),
  cl_data,
  most_specific = TRUE
)
utils::write.csv(sim_lin,
                 file.path(out_dir, "pbmc3k_cl_similarity_lin.csv"))
```

The most similar reviewed annotation pairs in the reference run were
biologically coherent:

| Label 1 | Label 2 | Lin similarity | Most specific common ancestor |
|---|---|---:|---|
| CD14+ Mono | CD16+ Mono | 0.911 | CD14-positive monocyte |
| Memory CD4 T | CD8 T | 0.758 | mature alpha-beta T cell |
| Naive CD4 T | Memory CD4 T | 0.712 | CD4-positive, alpha-beta T cell |
| B cell | NK | 0.689 | lymphocyte |
| CD14+ Mono | DC | 0.676 | mononuclear phagocyte |

## Roll annotations up to broader terms

Use `CLsuggestResolution()` to sweep candidate roll-up settings, but inspect the
actual mappings before locking a configuration. Matching the requested number
of groups is not, by itself, a biological selection criterion:

```{r rollup, purl = FALSE}
suggestion <- CLsuggestResolution(
  pbmc_terms$cl_id,
  cl_data,
  target_groups = 3,
  max_candidate_ancestor_count_values = 8:17,
  verbose = FALSE
)

# Several settings reach three groups. This mapping was selected after review
# and is locked for the case study rather than inferred again on every run.
rollup_min_group_size <- 3L
rollup_max_candidate_ancestor_count <- 8L

roll <- CLrollup(
  pbmc_terms$cl_id,
  cl_data,
  min_group_size = rollup_min_group_size,
  max_candidate_ancestor_count = rollup_max_candidate_ancestor_count,
  verbose = FALSE
)

roll_map <- roll$mapping
rollup_by_id <- setNames(roll_map$rolled_label, roll_map$original_id)
pbmc$cellontools_rollup_label <- unname(rollup_by_id[pbmc$cellontools_cl_id])
```

The selected roll-up compressed 9 reviewed terms to 3 broader groups:

| Original labels | Roll-up CL label |
|---|---|
| Naive CD4 T, Memory CD4 T, CD8 T, B, NK | lymphocyte |
| CD14+ Mono, FCGR3A+ Mono, DC | mononuclear leukocyte |
| Platelet | platelet |

This gives you two annotation layers: the reviewed fine-grained CL terms for
tables and the broader roll-up terms for overview plots or harmonisation across
datasets. The parameters and complete original-to-rolled mapping should be
exported together; rerunning only the target-group search is not a substitute
for preserving the reviewed decision.

## Compare marker enrichment to reviewed labels

Marker enrichment provides evidence for a cluster label. It should be interpreted
alongside the reviewed CL dictionary, not used as a mechanical replacement for
it. In this reference concordance analysis, markers are calculated for the
published Seurat annotation groups. That evaluates agreement with reviewed
labels; it is not an independent de novo annotation benchmark.

```{r markers, purl = FALSE}
marker_db <- CLmarkers("human")
data.frame(
  marker_rows = nrow(marker_db),
  unique_cl_terms = length(unique(marker_db$CL_ID)),
  unique_marker_symbols = length(unique(marker_db$marker_symbol))
)

all_markers <- FindAllMarkers(pbmc, only.pos = TRUE, min.pct = 0.25,
                              logfc.threshold = 0.25, verbose = FALSE)

# Seurat has used both names across releases.
lfc_col <- intersect(c("avg_log2FC", "avg_logFC"), colnames(all_markers))[1]
if (is.na(lfc_col)) {
  stop("Could not find a Seurat marker fold-change column.")
}

all_markers <- all_markers[order(all_markers$cluster,
                                 all_markers$p_val_adj,
                                 -all_markers[[lfc_col]]), ]

top_markers <- do.call(rbind, lapply(split(all_markers, all_markers$cluster),
                                     function(x) {
  x <- x[order(x$p_val_adj, -x[[lfc_col]]), ]
  utils::head(x, 25)
}))

gene_clusters <- split(top_markers$gene, top_markers$cluster)
# The published reference result demonstrates the package default background.
compare_res <- CLcompareCluster(gene_clusters,
                                geneType = "symbol",
                                species = "human",
                                universe = NULL,
                                pvalueCutoff = 0.05,
                                qvalueCutoff = 0.2)
compare_df <- as.data.frame(compare_res)

# For a study-specific analysis, use genes that could have been detected and
# treat the resulting table as a separate, explicitly labelled analysis.
assay_universe <- rownames(pbmc)
compare_res_assay_universe <- CLcompareCluster(
  gene_clusters,
  geneType = "symbol",
  species = "human",
  universe = assay_universe,
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2
)
compare_df_assay_universe <- as.data.frame(compare_res_assay_universe)
```

In the reference run, the human marker resource contained 140,337 marker rows
across 723 active CL terms and 15,474 unique marker symbols. `FindAllMarkers()`
returned 3,456 marker rows. `CLcompareCluster()` returned 674 enrichment rows
with the default marker-database universe and 663 rows with the 13,714 genes in
the pbmc3k assay as the background; all 9 groups had enrichment results under
both definitions.

Top marker-enriched terms compared with the reviewed CL labels:

| Seurat label | Top enriched term | Reviewed CL label | Exact ID match | Lin to reviewed |
|---|---|---|---|---:|
| Naive CD4 T | naive T cell | naive thymus-derived CD4-positive, alpha-beta T cell | FALSE | 0.911 |
| Memory CD4 T | naive T cell | CD4-positive, alpha-beta memory T cell | FALSE | 0.645 |
| CD14+ Mono | mononuclear phagocyte | CD14-positive monocyte | FALSE | 0.719 |
| B | B-2 B cell | B cell | FALSE | 0.790 |
| CD8 T | CD8-positive, alpha-beta memory T cell, CD45RO-positive | CD8-positive, alpha-beta T cell | FALSE | 0.851 |
| FCGR3A+ Mono | non-classical monocyte | CD14-low, CD16-positive monocyte | FALSE | 0.890 |
| NK | CD16-positive, CD56-dim natural killer cell, human | natural killer cell | FALSE | 0.781 |
| DC | CD1c-positive myeloid dendritic cell | dendritic cell | FALSE | 0.711 |
| Platelet | platelet | platelet | TRUE | 1.000 |

Only 1 of 9 top enriched terms was an exact CL-ID match, but the median Lin
similarity was 0.790 in the default-universe reference analysis. That is
expected: marker databases may point to a nearby subtype, parent, or sibling
term. Semantic similarity helps quantify whether the enrichment is biologically
close to the reviewed annotation. These numbers belong to the declared default
background; changing the universe changed some top-ranked terms, so the
concordance table must be regenerated for each background definition.

## From concordance to blinded label recovery

To evaluate marker-based label recovery, calculate markers from cluster IDs
without using the reference labels as identities. Lock one prediction per
cluster, then join `seurat_annotations` only for evaluation:

```{r blinded-label-recovery, purl = FALSE}
pbmc_blinded <- pbmc
Idents(pbmc_blinded) <- "seurat_clusters"

cluster_markers <- FindAllMarkers(
  pbmc_blinded,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25,
  verbose = FALSE
)

# Build top-marker lists by the numeric cluster IDs, run CLcompareCluster(),
# lock predictions, and only then attach the reference label table.
cluster_reference <- unique(pbmc_blinded[[]][
  , c("seurat_clusters", "seurat_annotations")
])
```

When several enriched terms are close, do not rank unrelated ontology branches
solely by `CLdepth()`: ancestor count measures specificity only within
ancestor-descendant-comparable terms. Prefer the best-supported term, use depth
to break ties within a lineage, or retain several candidates for review.

## Defensive reporting checklist

For reproducible scripts, export the intermediate tables as well as final
figures:

| Output | Why it helps |
|---|---|
| Reviewed label-to-CL dictionary | Makes annotation decisions auditable |
| OLS/search candidate table | Records evidence for shorthand labels |
| Cell metadata with original label, CL ID, label, and depth | Carries standardised terms into downstream analysis |
| Hierarchy nodes and edges | Makes ontology plots reproducible |
| Similarity matrix | Quantifies near-matches between terms |
| Roll-up parameters and mapping | Shows exactly which labels were collapsed and why |
| Marker table and declared universe | Defines the statistical input and background |
| `CLcompareCluster()` result table | Preserves marker evidence even when plots are filtered |
| Enrichment-vs-reviewed comparison | Separates exact matches from semantically close support |
| OBO release, URL, MD5, package versions, and `sessionInfo()` | Locks the computational provenance |
| Annotated Seurat RDS or metadata export | Preserves the reusable standardised annotation layer |

Guard network-dependent or optional branches:

```{r guards, purl = FALSE}
ols_candidates <- tryCatch(
  CLmap(pbmc_terms$query_label, max_results = 3,
        returnType = "all", verbose = FALSE),
  error = function(e) {
    data.frame(status = "api_error",
               error_message = conditionMessage(e))
  }
)

if (requireNamespace("clusterProfiler", quietly = TRUE)) {
  compare_res <- CLcompareCluster(gene_clusters,
                                  geneType = "symbol",
                                  species = "human",
                                  universe = rownames(pbmc))
  compare_df <- as.data.frame(compare_res)
  if (nrow(compare_df) == 0L) {
    message("No enriched CL terms passed the selected cutoffs.")
  }
}
```
## Save the reusable annotation layer and provenance

A manuscript result should be recoverable without rerunning network or
interactive steps. Save the annotated object, the ontology checksum, the
analysis background, package versions, and the complete R session:

```{r provenance, purl = FALSE}
run_provenance <- data.frame(
  generated_at = format(Sys.time(), tz = "UTC", usetz = TRUE),
  cellontools_version = as.character(packageVersion("CellOnTools")),
  seurat_version = as.character(packageVersion("Seurat")),

  ontology_release = attr(cl_data, "ontology_release"),
  ontology_source = attr(cl_data, "ontology_source"),
  ontology_md5 = unname(tools::md5sum(obo_file)),
  marker_universe = "declare the exact vector or default marker database",
  stringsAsFactors = FALSE
)

utils::write.csv(run_provenance,
                 file.path(out_dir, "run_provenance.csv"),
                 row.names = FALSE)
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
saveRDS(pbmc, file.path(out_dir, "pbmc3k_cellontools_annotated.rds"))
```
