---
title: "PBMC3K annotation workflow with CellOnTools"
author: "Di Wang"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{PBMC3K annotation workflow with CellOnTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Purpose

This vignette shows a manuscript-style workflow for using **CellOnTools** with a
single-cell object. It is based on a pbmc3k case study in which Seurat cluster
annotations were standardised to Cell Ontology (CL), written back to cell
metadata, rolled up to broader ontology terms, and checked against marker-gene
enrichment.

The code chunks are intentionally not evaluated during package checks because
they require SeuratData downloads, optional Bioconductor packages, plotting
packages, and network access for some mapping steps. Treat this vignette as a
template for your own analysis scripts.

```{r libraries}
library(CellOnTools)
library(Seurat)
library(SeuratData)
library(ggplot2)
```

Recommended optional packages for the full workflow:

| Step | Packages |
|---|---|
| Local CL loading | `ontologyIndex` |
| Semantic similarity | `ontologySimilarity` |
| Enrichment | `clusterProfiler`, `enrichplot` |
| Hierarchy plotting | `ggraph`, `igraph`, `ggplot2` |
| OLS candidate search | `rols` |

## Load the ontology and pbmc3k

For a manuscript or pipeline, prefer a local OBO file so that the ontology
release is explicit and rerunnable:

```{r load}
out_dir <- "pbmc3k_results"
dir.create(out_dir, showWarnings = FALSE)

obo_file <- file.path(out_dir, "cl.obo")
if (!file.exists(obo_file)) {
  CLdownload(dest_file = obo_file, overwrite = TRUE)
}
cl_data <- CLload(local_obo = obo_file, prefer_local = TRUE, verbose = FALSE)

pbmc_raw <- LoadData("pbmc3k")
pbmc <- LoadData("pbmc3k", type = "pbmc3k.final")
DefaultAssay(pbmc_raw) <- "RNA"
DefaultAssay(pbmc) <- "RNA"
Idents(pbmc) <- pbmc$seurat_annotations
```

The reference run behind this vignette used 2,700 raw pbmc3k cells and 2,638
processed `pbmc3k.final` cells.

## Build a reviewed CL dictionary

Start with a small table that keeps the original labels, search labels, and
reviewed CL labels separate. This makes the mapping auditable:

```{r reviewed-dictionary}
pbmc_terms <- data.frame(
  seurat_label = c("Naive CD4 T", "Memory CD4 T", "CD14+ Mono", "B", "CD8 T",
                   "FCGR3A+ Mono", "NK", "DC", "Platelet"),
  plot_label = c("Naive CD4 T", "Memory CD4 T", "CD14+ Mono", "B cell",
                 "CD8 T", "CD16+ Mono", "NK", "DC", "Platelet"),
  query_label = c("naive CD4-positive alpha-beta T cell",
                  "memory CD4-positive alpha-beta T cell",
                  "CD14-positive monocyte",
                  "B cell",
                  "CD8-positive alpha-beta T cell",
                  "CD16-positive monocyte",
                  "natural killer cell",
                  "dendritic cell",
                  "platelet"),
  reviewed_cl_label = c(
    "naive thymus-derived CD4-positive, alpha-beta T cell",
    "CD4-positive, alpha-beta memory T cell",
    "CD14-positive monocyte",
    "B cell",
    "CD8-positive, alpha-beta T cell",
    "CD14-low, CD16-positive monocyte",
    "natural killer cell",
    "dendritic cell",
    "platelet"
  )
)
```

Use local search and OLS candidates to audit shorthand labels before finalising
the reviewed dictionary:

```{r search-and-map}
search_hits <- CLsearchLabel(
  c("CD4-positive", "monocyte", "natural killer cell",
    "dendritic cell", "platelet"),
  cl_data,
  max_results = 8
)

ols_candidates <- tryCatch(
  CLmap(pbmc_terms$query_label, max_results = 3,
        returnType = "all", verbose = FALSE),
  error = function(e) {
    data.frame(error_message = conditionMessage(e))
  }
)
```

Then resolve the reviewed labels to CL IDs and add ontology metadata:

```{r resolve-labels}
pbmc_terms$cl_id <- CLlabel2id(pbmc_terms$reviewed_cl_label,
                               cl_data, strict = TRUE)
pbmc_terms$cl_label <- CLid2label(pbmc_terms$cl_id, cl_data)
pbmc_terms$cl_depth <- CLdepth(pbmc_terms$cl_id, cl_data)
pbmc_terms$n_ancestors <- lengths(CLancestors(pbmc_terms$cl_id, cl_data))
pbmc_terms$n_descendants <- lengths(CLdescendants(pbmc_terms$cl_id, cl_data))
```

The reference run resolved all 9 labels:

| Seurat label | Cells | CL ID | CL label | Depth |
|---|---:|---|---|---:|
| Naive CD4 T | 697 | `CL:0000895` | naive thymus-derived CD4-positive, alpha-beta T cell | 21 |
| Memory CD4 T | 483 | `CL:0000897` | CD4-positive, alpha-beta memory T cell | 21 |
| CD14+ Mono | 480 | `CL:0001054` | CD14-positive monocyte | 23 |
| B | 344 | `CL:0000236` | B cell | 16 |
| CD8 T | 271 | `CL:0000625` | CD8-positive, alpha-beta T cell | 19 |
| FCGR3A+ Mono | 162 | `CL:0002396` | CD14-low, CD16-positive monocyte | 25 |
| NK | 155 | `CL:0000623` | natural killer cell | 17 |
| DC | 32 | `CL:0000451` | dendritic cell | 19 |
| Platelet | 14 | `CL:0000233` | platelet | 15 |

## Write CL terms back to cell metadata

Keep both the original annotation and the CL-standardised annotation. This
preserves provenance and makes downstream plotting straightforward:

```{r metadata}
id_by_label <- setNames(pbmc_terms$cl_id, pbmc_terms$seurat_label)
label_by_id <- setNames(pbmc_terms$cl_label, pbmc_terms$cl_id)
plot_by_id <- setNames(pbmc_terms$plot_label, pbmc_terms$cl_id)
depth_by_id <- setNames(pbmc_terms$cl_depth, pbmc_terms$cl_id)

pbmc$cellontools_cl_id <- unname(id_by_label[as.character(pbmc$seurat_annotations)])
pbmc$cellontools_cl_label <- unname(label_by_id[pbmc$cellontools_cl_id])
pbmc$cellontools_cl_plot_label <- unname(plot_by_id[pbmc$cellontools_cl_id])
pbmc$cellontools_cl_depth <- unname(depth_by_id[pbmc$cellontools_cl_id])

DimPlot(pbmc, group.by = "cellontools_cl_plot_label",
        label = TRUE, repel = TRUE) +
  NoLegend()
```

## Explore ontology relationships

Once annotations are CL IDs, you can inspect shared ancestors, hierarchy
subgraphs, and semantic similarity:

```{r relationships}
hier <- CLhierarchy(pbmc_terms$cl_id, cl_data,
                    include_ancestors = TRUE,
                    max_ancestor_count = NULL)

CLhierarchyPlot(pbmc_terms$cl_id, cl_data,
                include_ancestors = TRUE,
                max_ancestor_count = NULL,
                show_id = FALSE)

ic <- ontologySimilarity::descendants_IC(cl_data)
sim_lin <- CLsimilarityMatrix(pbmc_terms$cl_id, clData = cl_data,
                              method = "lin",
                              information_content = ic,
                              verbose = FALSE)
```

The most similar reviewed annotation pairs in the reference run were
biologically coherent:

| Label 1 | Label 2 | Lin similarity | Most specific common ancestor |
|---|---|---:|---|
| CD14+ Mono | CD16+ Mono | 0.911 | CD14-positive monocyte |
| Memory CD4 T | CD8 T | 0.758 | mature alpha-beta T cell |
| Naive CD4 T | Memory CD4 T | 0.712 | CD4-positive, alpha-beta T cell |
| B cell | NK | 0.689 | lymphocyte |
| CD14+ Mono | DC | 0.676 | mononuclear phagocyte |

## Roll annotations up to broader terms

Use `CLsuggestResolution()` to sweep candidate roll-up settings, then apply the
chosen setting with `CLrollup()`:

```{r rollup}
suggestion <- CLsuggestResolution(
  pbmc_terms$cl_id,
  cl_data,
  target_groups = 3,
  max_ancestor_count_values = c(16L, 17L, 18L, 19L, 20L, 21L, 22L),
  verbose = FALSE
)

roll <- CLrollup(pbmc_terms$cl_id, cl_data,
                 min_group_size = 2L,
                 max_ancestor_count = 17L,
                 verbose = FALSE)

roll_map <- roll$mapping
rollup_by_id <- setNames(roll_map$rolled_label, roll_map$original_id)
pbmc$cellontools_rollup_label <- unname(rollup_by_id[pbmc$cellontools_cl_id])
```

The selected roll-up compressed 9 reviewed terms to 3 broader groups:

| Original labels | Roll-up CL label |
|---|---|
| Naive CD4 T, Memory CD4 T, B, CD8 T, NK | lymphocyte |
| CD14+ Mono, FCGR3A+ Mono, DC | mononuclear phagocyte |
| Platelet | platelet |

This gives you two annotation layers: the reviewed fine-grained CL terms for
tables and the broader roll-up terms for overview plots or harmonisation across
datasets.

## Compare marker enrichment to reviewed labels

Marker enrichment provides evidence for a cluster label. It should be interpreted
alongside the reviewed CL dictionary, not used as a mechanical replacement for
it.

```{r markers}
marker_db <- CLmarkers("human")
data.frame(
  marker_rows = nrow(marker_db),
  unique_cl_terms = length(unique(marker_db$CL_ID)),
  unique_marker_symbols = length(unique(marker_db$marker_symbol))
)

all_markers <- FindAllMarkers(pbmc, only.pos = TRUE, min.pct = 0.25,
                              logfc.threshold = 0.25, verbose = FALSE)

# Seurat has used both names across releases.
lfc_col <- intersect(c("avg_log2FC", "avg_logFC"), colnames(all_markers))[1]
if (is.na(lfc_col)) {
  stop("Could not find a Seurat marker fold-change column.")
}

all_markers <- all_markers[order(all_markers$cluster,
                                 all_markers$p_val_adj,
                                 -all_markers[[lfc_col]]), ]

top_markers <- do.call(rbind, lapply(split(all_markers, all_markers$cluster),
                                     function(x) {
  x <- x[order(x$p_val_adj, -x[[lfc_col]]), ]
  utils::head(x, 25)
}))

gene_clusters <- split(top_markers$gene, top_markers$cluster)
compare_res <- CLcompareCluster(gene_clusters,
                                geneType = "symbol",
                                species = "human",
                                pvalueCutoff = 0.05,
                                qvalueCutoff = 0.2)
compare_df <- as.data.frame(compare_res)
```

In the reference run, the human marker resource contained 140,337 marker rows
across 728 CL terms and 15,474 unique marker symbols. `FindAllMarkers()`
returned 3,456 marker rows; `CLcompareCluster()` returned 674 enrichment rows.

Top marker-enriched terms compared with the reviewed CL labels:

| Seurat label | Top enriched term | Reviewed CL label | Exact ID match | Lin to reviewed |
|---|---|---|---|---:|
| Naive CD4 T | naive T cell | naive thymus-derived CD4-positive, alpha-beta T cell | FALSE | 0.911 |
| Memory CD4 T | naive T cell | CD4-positive, alpha-beta memory T cell | FALSE | 0.645 |
| CD14+ Mono | mononuclear phagocyte | CD14-positive monocyte | FALSE | 0.719 |
| B | B-2 B cell | B cell | FALSE | 0.790 |
| CD8 T | CD8-positive, alpha-beta memory T cell, CD45RO-positive | CD8-positive, alpha-beta T cell | FALSE | 0.851 |
| FCGR3A+ Mono | non-classical monocyte | CD14-low, CD16-positive monocyte | FALSE | 0.890 |
| NK | CD16-positive, CD56-dim natural killer cell, human | natural killer cell | FALSE | 0.781 |
| DC | CD1c-positive myeloid dendritic cell | dendritic cell | FALSE | 0.711 |
| Platelet | platelet | platelet | TRUE | 1.000 |

Only 1 of 9 top enriched terms was an exact CL-ID match, but the median Lin
similarity was 0.790. That is expected: marker databases may point to a nearby
subtype, parent, or sibling term. Semantic similarity helps quantify whether
the enrichment is biologically close to the reviewed annotation.

## Defensive reporting checklist

For reproducible scripts, export the intermediate tables as well as final
figures:

| Output | Why it helps |
|---|---|
| Reviewed label-to-CL dictionary | Makes annotation decisions auditable |
| OLS/search candidate table | Records evidence for shorthand labels |
| Cell metadata with CL ID, label, and depth | Carries standardised terms into downstream analysis |
| Hierarchy nodes and edges | Makes ontology plots reproducible |
| Similarity matrix | Quantifies near-matches between terms |
| Roll-up mapping | Shows exactly which labels were collapsed |
| `CLcompareCluster()` result table | Preserves marker evidence even when plots are filtered |
| Enrichment-vs-reviewed comparison | Separates exact matches from semantically close support |

Guard network-dependent or optional branches:

```{r guards}
ols_candidates <- tryCatch(
  CLmap(pbmc_terms$query_label, max_results = 3,
        returnType = "all", verbose = FALSE),
  error = function(e) {
    data.frame(status = "api_error",
               error_message = conditionMessage(e))
  }
)

if (requireNamespace("clusterProfiler", quietly = TRUE)) {
  compare_res <- CLcompareCluster(gene_clusters,
                                  geneType = "symbol",
                                  species = "human")
  compare_df <- as.data.frame(compare_res)
  if (nrow(compare_df) == 0L) {
    message("No enriched CL terms passed the selected cutoffs.")
  }
}
```
