---
title: "PBMC3K annotation workflow with CellOnTools"
author: "Di Wang"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{PBMC3K annotation workflow with CellOnTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Purpose

This vignette shows a manuscript-style workflow for using **CellOnTools** with a
single-cell object. It is based on a pbmc3k case study in which published Seurat
annotations were standardised to Cell Ontology (CL), written back to cell
metadata, rolled up to broader ontology terms, and compared with marker-gene
enrichment evidence.

The raw 2,700-cell object is used only to show the standard Seurat QC context;
all annotation analyses use the separately distributed, published 2,638-cell
`pbmc3k.final` object. The displayed QC thresholds are not applied to create
that object in the standalone workflow. Accordingly, this is a standardisation
and reference-conditioned concordance case study, not an end-to-end benchmark
of preprocessing, clustering, or independent automated cell typing.

The full workflow chunks are intentionally not evaluated during package checks
because they require external pbmc3k objects, optional Bioconductor and plotting
packages, and network or console interaction for candidate-search examples.
The code and reported reference values below follow the standalone analysis
script, `pbmc3k/pbmc3k_cellontools_case_study.R`.

The following small offline check is evaluated when the vignette is built, so
the document still verifies its CellOnTools installation without downloading
data or contacting an ontology service:

```{r offline-smoke, eval = TRUE, purl = TRUE}
human_markers <- CellOnTools::CLmarkers("human")
stopifnot(
  is.data.frame(human_markers),
  all(c("CL_ID", "marker_symbol") %in% names(human_markers))
)
```

```{r libraries, purl = FALSE}
library(CellOnTools)
library(Seurat)
library(SeuratData)
library(ggplot2)

options(stringsAsFactors = FALSE)
set.seed(42)
```

Recommended optional packages for the full workflow:

| Step | Packages |
|---|---|
| Single-cell object | `Seurat` |
| Local CL loading | `ontologyIndex` |
| Semantic similarity | `ontologySimilarity` |
| Enrichment | `clusterProfiler`, `enrichplot` |
| Hierarchy plotting | `ggraph`, `igraph`, `ggplot2` |
| OLS candidate search | `rols` |

## Load the ontology and pbmc3k

The standalone script creates separate table and figure directories and keeps a
local OBO file with the results:

```{r load, purl = FALSE}
script_dir <- local({
  args <- grep("^--file=", commandArgs(FALSE), value = TRUE)
  if (length(args)) {
    dirname(normalizePath(sub("^--file=", "", args[1]), winslash = "/"))
  } else if (!is.null(sys.frames()[[1]]$ofile)) {
    dirname(normalizePath(sys.frames()[[1]]$ofile, winslash = "/"))
  } else {
    getwd()
  }
})

out_dir <- file.path(script_dir, "pbmc3k_results")
table_dir <- file.path(out_dir, "tables")
fig_dir <- file.path(out_dir, "figures")
for (d in c(table_dir, fig_dir)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}

save_tab <- function(x, file, row.names = FALSE) {
  utils::write.csv(x, file.path(table_dir, file), row.names = row.names)
}
save_fig <- function(plot, file, width = 8, height = 6, dir = fig_dir) {
  ggsave(file.path(dir, file), plot, width = width, height = height,
         dpi = 300, bg = "white")
}
wrap_label <- function(x, width = 24) {
  vapply(x, function(label) {
    if (is.na(label) || !nzchar(label)) return(NA_character_)
    paste(strwrap(label, width = width), collapse = "\n")
  }, character(1))
}

obo_file <- file.path(out_dir, "cl.obo")
if (!file.exists(obo_file)) {
  CLdownload(dest_file = obo_file, overwrite = TRUE)
}
cl_data <- CLload(local_obo = obo_file, prefer_local = TRUE, verbose = FALSE)

pbmc_raw <- LoadData("pbmc3k")
DefaultAssay(pbmc_raw) <- "RNA"
pbmc_raw[["percent.mt"]] <- PercentageFeatureSet(pbmc_raw, pattern = "^MT-")

qc_df <- rbind(
  data.frame(metric = "Detected genes", value = pbmc_raw$nFeature_RNA),
  data.frame(metric = "UMIs", value = pbmc_raw$nCount_RNA),
  data.frame(metric = "% mitochondrial", value = pbmc_raw$percent.mt)
)
qc_cuts <- data.frame(
  metric = c("Detected genes", "Detected genes", "% mitochondrial"),
  cut = c(200, 2500, 5)
)
save_fig(
  ggplot(qc_df, aes(value)) +
    geom_histogram(bins = 200, fill = "#4C78A8", colour = "white",
                   linewidth = 0.1) +
    geom_vline(data = qc_cuts, aes(xintercept = cut),
               colour = "#D62728", linewidth = 0.5, linetype = "dashed") +
    facet_wrap(~ metric, scales = "free", ncol = 1) +
    labs(x = NULL, y = "Cells",
         title = "pbmc3k QC with Seurat tutorial thresholds") +
    theme_classic(base_size = 11),
  "pbmc3k_qc_distributions.pdf", width = 6, height = 7
)

# The thresholds above are shown for QC context; they are not used to subset
# pbmc_raw in this workflow.
pbmc_final <- LoadData("pbmc3k", type = "pbmc3k.final")
DefaultAssay(pbmc_final) <- "RNA"

# Published reference labels define all downstream marker groups.
Idents(pbmc_final) <- pbmc_final$seurat_annotations
if (!"percent.mt" %in% colnames(pbmc_final[[]])) {
  pbmc_final[["percent.mt"]] <- PercentageFeatureSet(pbmc_final, pattern = "^MT-")
}

cell_counts <- as.data.frame(table(pbmc_final$seurat_annotations))
colnames(cell_counts) <- c("seurat_label", "n_cells")
cell_counts$seurat_label <- as.character(cell_counts$seurat_label)
cell_counts <- cell_counts[order(-cell_counts$n_cells), ]
save_tab(cell_counts, "pbmc3k_seurat_annotation_counts.csv")

save_fig(
  DimPlot(pbmc_final, group.by = "seurat_annotations",
          label = TRUE, repel = TRUE) +
    ggtitle("pbmc3k Seurat annotations") +
    theme_classic(base_size = 11) +
    theme(legend.position = "none"),
  "pbmc3k_umap_seurat_annotations.pdf", width = 5, height = 5
)
```

The raw object contains 2,700 cells, whereas the separately distributed
`pbmc3k.final` object contains 2,638 cells. The standalone script does not
calculate a QC-pass count or derive the latter object by subsetting the former.

## Build a reviewed CL dictionary

The online and interactive search calls in the standalone script are
exploratory evidence only: their return values do not feed the downstream
analysis. The reviewed CL IDs are locked explicitly so that the deterministic
analysis does not depend on a live OLS response or a console selection:

```{r search-and-map, purl = FALSE}
seurat_labels <- sort(unique(as.character(pbmc_final$seurat_annotations)))
CLsearchLabel(seurat_labels, cl_data, exact_match = TRUE)

CLmap("CD14+ Mono", returnType = "all", max_results = 10, verbose = FALSE)
CLmap("CD8 T", returnType = "all", max_results = 10, verbose = FALSE)
CLmap("Memory CD4 T", returnType = "all", max_results = 10, verbose = FALSE)
CLmap("Naive CD4 T", returnType = "all", max_results = 10, verbose = FALSE)

CLmapInteractive("B", returnType = "all", n_candidates = 10)
CLmapInteractive("DC", returnType = "all", n_candidates = 10)
CLmapInteractive("FCGR3A+ Mono", returnType = "all", n_candidates = 10)
CLmapInteractive("NK", returnType = "all", n_candidates = 10)
```

`CLmapInteractive()` is intended for console review. In a non-interactive
session it may automatically accept a top-ranked candidate, so these calls
should not be treated as the source of the locked downstream dictionary.

```{r reviewed-dictionary, purl = FALSE}
pbmc_terms <- data.frame(
  seurat_label = c(
    "B", "CD14+ Mono", "CD8 T", "DC", "FCGR3A+ Mono",
    "Memory CD4 T", "Naive CD4 T", "NK", "Platelet"
  ),
  query_label = c(
    "B cell", "CD14+ Mono", "CD8 T", "dendritic cell", "CD16+ mono",
    "Memory CD4 T", "Naive CD4 T", "natural killer cell", "Platelet"
  ),
  cl_id = c(
    "CL:0000236", "CL:0001054", "CL:0000625", "CL:0000451",
    "CL:0002396", "CL:0000897", "CL:0000895", "CL:0000623",
    "CL:0000233"
  ),
  standardization_method = c(
    "CLmapInteractive", "CLmap", "CLmap", "CLmapInteractive",
    "CLmapInteractive", "CLmap", "CLmap", "CLmapInteractive",
    "CLsearchLabel"
  ),
  stringsAsFactors = FALSE
)

pbmc_terms$cl_label <- unname(
  CLid2label(pbmc_terms$cl_id, cl_data, strict = TRUE)
)
pbmc_terms$cl_depth <- unname(CLdepth(pbmc_terms$cl_id, cl_data))

anc <- CLancestors(pbmc_terms$cl_id, cl_data)
des <- CLdescendants(pbmc_terms$cl_id, cl_data)
pbmc_terms$n_ancestors <- lengths(anc)[pbmc_terms$cl_id]
pbmc_terms$n_descendants <- lengths(des)[pbmc_terms$cl_id]
pbmc_terms$n_cells <- cell_counts$n_cells[
  match(pbmc_terms$seurat_label, cell_counts$seurat_label)
]

pbmc_terms <- pbmc_terms[, c(
  "seurat_label", "n_cells", "query_label", "standardization_method",
  "cl_id", "cl_label", "cl_depth", "n_ancestors", "n_descendants"
)]
pbmc_terms$ontology_release <- attr(cl_data, "ontology_release")
save_tab(pbmc_terms, "pbmc3k_reviewed_cl_dictionary.csv")
```

The locked case-study dictionary contains:

| Seurat label | Cells | CL ID | CL label | CL-only ancestor count |
|---|---:|---|---|---:|
| B | 344 | `CL:0000236` | B cell | 10 |
| CD14+ Mono | 480 | `CL:0001054` | CD14-positive monocyte | 17 |
| CD8 T | 271 | `CL:0000625` | CD8-positive, alpha-beta T cell | 13 |
| DC | 32 | `CL:0000451` | dendritic cell | 13 |
| FCGR3A+ Mono | 162 | `CL:0002396` | CD14-low, CD16-positive monocyte | 19 |
| Memory CD4 T | 483 | `CL:0000897` | CD4-positive, alpha-beta memory T cell | 15 |
| Naive CD4 T | 697 | `CL:0000895` | naive thymus-derived CD4-positive, alpha-beta T cell | 15 |
| NK | 155 | `CL:0000623` | natural killer cell | 11 |
| Platelet | 14 | `CL:0000233` | platelet | 9 |

`CLdepth()` is the number of distinct proper ancestors within the CL
namespace. It is not a graph-hop distance, and values from unrelated ontology
branches should not be treated as a universal biological scale.

## Write CL terms back to cell metadata

The standalone script writes the reviewed CL ID and canonical CL label to each
cell while preserving the original Seurat annotation:

```{r metadata, purl = FALSE}
id_by_label <- setNames(pbmc_terms$cl_id, pbmc_terms$seurat_label)
cl_by_id <- setNames(pbmc_terms$cl_label, pbmc_terms$cl_id)

pbmc_final$cellontools_cl_id <- unname(
  id_by_label[as.character(pbmc_final$seurat_annotations)]
)
pbmc_final$cellontools_cl_label <- unname(
  cl_by_id[pbmc_final$cellontools_cl_id]
)

save_fig(
  DimPlot(pbmc_final, group.by = "cellontools_cl_label",
          label = TRUE, repel = TRUE, label.size = 3) +
    ggtitle("pbmc3k standardized to Cell Ontology") +
    theme_classic(base_size = 11) +
    theme(legend.position = "none"),
  "pbmc3k_umap_cell_ontology.pdf", width = 5, height = 5
)
```

The script does not write CL depth or a separate plot label to cell metadata.

## Explore ontology relationships

The standalone script exports the hierarchy induced by the reviewed terms and
all their ancestors, then calculates a Lin semantic-similarity matrix:

```{r relationships, purl = FALSE}
hier <- CLhierarchy(
  pbmc_terms$cl_id,
  cl_data,
  include_ancestors = TRUE,
  max_hops = NULL
)
save_tab(hier$nodes, "pbmc3k_cl_hierarchy_nodes.csv")
save_tab(hier$edges, "pbmc3k_cl_hierarchy_edges.csv")

save_fig(
  CLhierarchyPlot(
    pbmc_terms$cl_id,
    cl_data,
    include_ancestors = TRUE,
    max_hops = NULL,
    show_id = TRUE,
    label_size = 3,
    node_size = 2,
    query_color = "#D62728",
    ancestor_color = "#4C78A8"
  ),
  "pbmc3k_cell_ontology_hierarchy.pdf", width = 15, height = 10
)

plot_pbmc_hierarchy_publication <- function(pbmc_terms, cl_data) {
  hierarchy <- CLhierarchy(
    pbmc_terms$cl_id, cl_data, include_ancestors = TRUE, max_hops = NULL
  )
  biological_backbone <- c(
    "CL:0000988", "CL:0000081", "CL:0000225",
    "CL:0000763", "CL:0000234", "CL:0000518"
  )
  nodes <- subset(
    hierarchy$nodes,
    grepl("^CL:", id) & !is.na(label) &
      (is_query | ancestor_count >= 11 | id %in% biological_backbone)
  )
  nodes$plot_label <- ifelse(
    nodes$is_query,
    paste0(wrap_label(nodes$label, 34), "\n", nodes$id),
    wrap_label(nodes$label, 30)
  )
  nodes$node_class <- factor(
    ifelse(nodes$is_query, "Query terms", "Ontology ancestors"),
    levels = c("Ontology ancestors", "Query terms")
  )
  edges <- subset(hierarchy$edges, from %in% nodes$id & to %in% nodes$id)
  edges <- data.frame(from = edges$to, to = edges$from, stringsAsFactors = FALSE)
  graph <- igraph::graph_from_data_frame(
    edges, vertices = nodes, directed = TRUE
  )
  ggraph::ggraph(graph, layout = "sugiyama") +
    ggraph::geom_edge_diagonal(
      colour = "#B8B8B8", linewidth = 0.35, alpha = 0.85,
      arrow = grid::arrow(length = grid::unit(1.4, "mm"), type = "closed"),
      end_cap = ggraph::circle(2.2, "mm")
    ) +
    ggraph::geom_node_point(
      aes(fill = node_class, size = node_class),
      shape = 21, colour = "white", stroke = 0.35
    ) +
    ggraph::geom_node_label(
      aes(label = plot_label, fill = node_class),
      size = 2.25, label.size = 0.18, colour = "#1F2933",
      lineheight = 0.92, repel = TRUE, max.overlaps = Inf,
      box.padding = 0.30, point.padding = 0.18, min.segment.length = 0,
      show.legend = FALSE
    ) +
    scale_fill_manual(values = c(
      "Ontology ancestors" = "#E8EEF5", "Query terms" = "#D9483B"
    )) +
    scale_size_manual(values = c(
      "Ontology ancestors" = 2.2, "Query terms" = 3.4
    )) +
    guides(size = "none", fill = guide_legend(
      override.aes = list(shape = 21, size = 4, colour = "white")
    )) +
    labs(title = NULL, subtitle = NULL, fill = NULL) +
    theme_void(base_size = 8) +
    theme(legend.position = "bottom")
}

p_hierarchy <- plot_pbmc_hierarchy_publication(pbmc_terms, cl_data)
save_fig(
  p_hierarchy, "pbmc3k_cell_ontology_hierarchy_pub.pdf",
  width = 9.2, height = 6.4
)

ic <- ontologySimilarity::descendants_IC(cl_data)
sim_lin <- CLsimilarityMatrix(
  pbmc_terms$cl_id,
  clData = cl_data,
  method = "lin",
  information_content = ic,
  verbose = FALSE
)
save_tab(sim_lin, "pbmc3k_cl_similarity_lin.csv", row.names = TRUE)
save_fig(
  pheatmap::pheatmap(
    mat = sim_lin,
    color = grDevices::colorRampPalette(c("#F7FBFF", "#08306B"))(100),
    border_color = "white",
    treeheight_row = 3,
    treeheight_col = 3,
    show_colnames = FALSE
  ),
  "pbmc3k_cl_similarity_heatmap.pdf", width = 8, height = 3.5
)

# A single Resnik comparison is printed to the console.
CLsimilarity(
  "CL:0001054", "CL:0002396", cl_data,
  method = "resnik", information_content = ic
)
```

Selected biologically coherent pairs in the locked Lin matrix are:

| Label 1 | Label 2 | Lin similarity |
|---|---|---:|
| CD14+ Mono | CD16+ Mono | 0.911135 |
| Memory CD4 T | CD8 T | 0.758919 |
| Naive CD4 T | Memory CD4 T | 0.713261 |
| B cell | NK | 0.690498 |
| CD14+ Mono | DC | 0.677934 |

The information content is derived from ontology descendant structure, not PBMC
expression or cell frequencies. Lin similarity is therefore an
ontology-semantic measure, not an expression correlation or accuracy statistic.
The standalone script does not call `CLcommonAncestor()`; shared ancestors are
represented implicitly in the hierarchy.

## Roll annotations up to broader terms

`CLsuggestResolution()` scans candidate parameters, but the standalone case
study does not automatically accept `suggestion$best`. It locks a reviewed
configuration after inspecting candidate mappings:

```{r rollup, purl = FALSE}
resolution_suggestion <- CLsuggestResolution(
  pbmc_terms$cl_id,
  cl_data,
  target_groups = 3L,
  max_candidate_ancestor_count_values = 8:17,
  verbose = FALSE
)

resolution_scan <- do.call(rbind, lapply(
  resolution_suggestion$all_candidates,
  function(candidate) {
    data.frame(
      min_group_size = candidate$min_group_size,
      max_candidate_ancestor_count = if (
        is.null(candidate$max_candidate_ancestor_count)
      ) NA_integer_ else candidate$max_candidate_ancestor_count,
      n_groups = candidate$n_groups,
      n_rolled = candidate$n_rolled,
      compression_ratio = candidate$compression_ratio,
      deviation = candidate$deviation,
      error = if (is.null(candidate$error)) NA_character_ else candidate$error,
      stringsAsFactors = FALSE
    )
  }
))
save_tab(resolution_scan, "pbmc3k_rollup_resolution_scan.csv")

rollup_min_group_size <- 3L
rollup_max_candidate_ancestor_count <- 8L

inspection_configs <- data.frame(
  min_group_size = c(3L, 2L, 3L),
  max_candidate_ancestor_count = c(8L, 11L, 17L)
)
rollup_inspection <- do.call(rbind, lapply(
  seq_len(nrow(inspection_configs)),
  function(i) {
    config <- inspection_configs[i, ]
    result <- CLrollup(
      pbmc_terms$cl_id,
      cl_data,
      min_group_size = config$min_group_size,
      max_candidate_ancestor_count = config$max_candidate_ancestor_count,
      verbose = FALSE
    )
    output <- result$mapping
    output$min_group_size <- config$min_group_size
    output$max_candidate_ancestor_count <- config$max_candidate_ancestor_count
    output
  }
))
save_tab(rollup_inspection, "pbmc3k_rollup_mapping_inspection.csv")

rollup_result <- CLrollup(
  pbmc_terms$cl_id,
  cl_data,
  min_group_size = rollup_min_group_size,
  max_candidate_ancestor_count = rollup_max_candidate_ancestor_count,
  verbose = FALSE
)
rollup_mapping <- rollup_result$mapping
rollup_mapping$seurat_label <- pbmc_terms$seurat_label[
  match(rollup_mapping$original_id, pbmc_terms$cl_id)
]
rollup_mapping$n_cells <- pbmc_terms$n_cells[
  match(rollup_mapping$original_id, pbmc_terms$cl_id)
]
rollup_mapping <- rollup_mapping[, c(
  "seurat_label", "n_cells", "original_id", "original_label",
  "rolled_id", "rolled_label", "was_rolled"
)]
save_tab(rollup_mapping, "pbmc3k_cell_ontology_rollup.csv")

rollup_by_id <- setNames(
  rollup_mapping$rolled_label,
  rollup_mapping$original_id
)
pbmc_final$cellontools_rollup_label <- unname(
  rollup_by_id[pbmc_final$cellontools_cl_id]
)
n_rollup_groups <- length(unique(pbmc_final$cellontools_rollup_label))
save_fig(
  DimPlot(pbmc_final, group.by = "cellontools_rollup_label",
          label = TRUE, repel = TRUE) +
    ggtitle(sprintf("CellOnTools roll-up (%d groups)", n_rollup_groups)) +
    theme_classic(base_size = 11) +
    theme(legend.position = "none"),
  "pbmc3k_umap_cell_ontology_rollup.pdf", width = 5, height = 5
)
```

The selected configuration compresses the nine reviewed terms to three groups:

| Original labels | Roll-up CL label |
|---|---|
| Naive CD4 T, Memory CD4 T, CD8 T, B, NK | lymphocyte |
| CD14+ Mono, FCGR3A+ Mono, DC | mononuclear leukocyte |
| Platelet | platelet |

`min_group_size` counts distinct input CL terms, not cells. The candidate cap
is a CL ancestor-count specificity limit, not a maximum hop distance.

## Compare marker enrichment to reviewed labels

Markers are calculated after setting identities to the published Seurat
annotations. This is reference-conditioned evidence for the reviewed labels,
not a blinded marker-based annotation benchmark:

```{r markers, purl = FALSE}
marker_db <- CLmarkers("human", check_unique = FALSE)

all_markers <- FindAllMarkers(
  pbmc_final,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25,
  verbose = FALSE
)

lfc_col <- intersect(c("avg_log2FC", "avg_logFC"), colnames(all_markers))[1]
if (is.na(lfc_col)) {
  stop("Could not find a Seurat marker fold-change column.")
}
all_markers <- all_markers[
  order(all_markers$cluster, all_markers$p_val_adj,
        -all_markers[[lfc_col]]),
]
top_markers <- do.call(rbind, lapply(
  split(all_markers, all_markers$cluster),
  function(x) {
    x <- x[order(x$p_val_adj, -x[[lfc_col]]), ]
    utils::head(x, 25)
  }
))
save_tab(all_markers, "pbmc3k_all_seurat_markers.csv")
save_tab(top_markers, "pbmc3k_top25_seurat_markers.csv")

gene_clusters <- split(top_markers$gene, top_markers$cluster)
compare_res <- CLcompareCluster(
  gene_clusters,
  geneType = "symbol",
  species = "human",
  universe = NULL,
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2
)
compare_df <- as.data.frame(compare_res)
save_tab(compare_df, "pbmc3k_marker_enrichment_by_annotation.csv")

if (nrow(compare_df) == 0L) {
  stop("CLcompareCluster returned no enriched Cell Ontology terms.")
}

top_terms <- do.call(rbind, lapply(
  split(compare_df, compare_df$Cluster),
  function(x) {
    x <- x[order(x$p.adjust, x$pvalue), ]
    utils::head(x[, c("Cluster", "ID", "Description", "p.adjust")], 3)
  }
))
save_tab(top_terms, "pbmc3k_top_marker_enriched_cl_terms.csv")
save_fig(
  enrichplot::dotplot(compare_res, showCategory = 5, font.size = 9) +
    scale_y_discrete(labels = function(x) wrap_label(x, 50)) +
    ggtitle("CellOnTools marker enrichment across pbmc3k annotations") +
    theme(
      plot.title = element_text(hjust = 0.5),
      axis.text.x = element_text(angle = 40, hjust = 1)
    ),
  "pbmc3k_marker_enrichment_dotplot.pdf", width = 11, height = 12
)

b_enr <- CLenricher(
  gene_clusters[["B"]],
  geneType = "symbol",
  species = "human",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2
)
b_enr_df <- if (is.null(b_enr)) data.frame() else as.data.frame(b_enr)
save_tab(b_enr_df, "pbmc3k_Bcell_marker_enrichment.csv")
save_fig(
  enrichplot::dotplot(b_enr),
  "pbmc3k_Bcell_marker_enrichment_dotplot.pdf", width = 6.5, height = 4
)
```

The locked reference run produced 3,456 Seurat marker rows, 225 top-marker rows,
and 674 enrichment rows with `universe = NULL`. Here the default universe is
the set of genes in the bundled marker database. The standalone script mentions
an assay-gene universe only as a possible separately labelled sensitivity
analysis; it does not run or report that analysis.

## Exploratory marker-to-CL concordance

The standalone script does not simply select the smallest adjusted P value. It
retains candidates within ten-fold of the minimum adjusted P value, then selects
the candidate with the largest CL-only ancestor count, breaking remaining ties
by adjusted and raw P values:

```{r marker-concordance, purl = FALSE}
assign_cluster <- function(df) {
  df <- df[order(df$p.adjust, df$pvalue), ]
  cand <- df[
    df$p.adjust <= min(df$p.adjust, na.rm = TRUE) * 10,
    , drop = FALSE
  ]
  cand$cl_depth <- as.integer(CLdepth(cand$ID, cl_data))
  cand$cl_depth[is.na(cand$cl_depth)] <- -1L
  cand <- cand[order(-cand$cl_depth, cand$p.adjust, cand$pvalue), ]
  cand[1, c("Cluster", "ID", "Description", "p.adjust", "cl_depth"),
       drop = FALSE]
}

annotation <- do.call(
  rbind,
  lapply(split(compare_df, compare_df$Cluster), assign_cluster)
)
annotation$seurat_label <- as.character(annotation$Cluster)
m <- match(annotation$seurat_label, pbmc_terms$seurat_label)
annotation$n_cells <- pbmc_terms$n_cells[m]
annotation$standardized_id <- pbmc_terms$cl_id[m]
annotation$standardized_label <- pbmc_terms$cl_label[m]
annotation$lin_to_standardized <- round(mapply(function(a, b) {
  if (is.na(a) || is.na(b) || !a %in% cl_data$id || !b %in% cl_data$id) {
    return(NA_real_)
  }
  CLsimilarity(a, b, cl_data, method = "lin", information_content = ic)
}, annotation$ID, annotation$standardized_id), 3)

annotation$same_lineage <- mapply(function(a, b) {
  if (is.na(a) || is.na(b) || !a %in% cl_data$id || !b %in% cl_data$id) {
    return(NA)
  }
  a_anc <- CLancestors(a, cl_data)[[1]]
  b_anc <- CLancestors(b, cl_data)[[1]]
  a == b || a %in% b_anc || b %in% a_anc
}, annotation$ID, annotation$standardized_id)

annotation <- annotation[order(m), c(
  "seurat_label", "n_cells", "ID", "Description", "p.adjust", "cl_depth",
  "standardized_id", "standardized_label", "lin_to_standardized",
  "same_lineage"
)]
names(annotation)[3:4] <- c("cl_id", "cl_term")
save_tab(annotation, "pbmc3k_de_novo_marker_cl_annotation.csv")

plot_df <- annotation
plot_df$neg_log10_padj <- -log10(plot_df$p.adjust)
save_fig(
  ggplot(
    plot_df,
    aes(
      x = standardized_label, y = cl_term, size = neg_log10_padj,
      colour = lin_to_standardized, shape = same_lineage
    )
  ) +
    geom_point(alpha = 0.9) +
    scale_colour_viridis_c(option = "C", limits = c(0, 1)) +
    labs(
      x = "Standardized CL annotation",
      y = "Marker-enriched CL term",
      size = "-log10(adj. P)",
      colour = "Lin similarity",
      shape = "Same lineage",
      title = "Marker-based Cell Ontology concordance"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5, face = "bold")
    ),
  "pbmc3k_marker_cl_annotation_concordance.pdf", width = 7.5, height = 6
)
```

The current standalone-script result is:

| Seurat label | Selected marker-enriched term | Lin to reviewed | Same lineage |
|---|---|---:|:---:|
| B | B-2 B cell | 0.791 | TRUE |
| CD14+ Mono | CD14-positive, CD16-positive monocyte | 0.947 | TRUE |
| CD8 T | CD8-positive, alpha-beta memory T cell, CD45RO-positive | 0.851 | TRUE |
| DC | CD1c-positive myeloid dendritic cell | 0.712 | TRUE |
| FCGR3A+ Mono | non-classical monocyte | 0.891 | TRUE |
| Memory CD4 T | effector memory CD4-positive, alpha-beta T cell | 0.875 | TRUE |
| Naive CD4 T | naive T cell | 0.911 | TRUE |
| NK | CD16-positive, CD56-dim natural killer cell, human | 0.782 | TRUE |
| Platelet | platelet | 1.000 | TRUE |

This gives one exact CL-ID match in nine groups, nine ancestor/descendant-
comparable pairs, and a median Lin similarity of 0.875. The output filename is
retained to match the standalone script, but the result is not an independent de
novo prediction because its marker groups were defined by the published labels.

Ancestor count is not globally comparable across unrelated ontology branches.
The implemented depth-prioritised rule is an exploratory heuristic. A blinded
benchmark should use cluster IDs, lock predictions before joining reference
labels, and restrict any depth-based tie-break to biologically comparable
candidates.

## From concordance to blinded label recovery

A separate label-recovery evaluation would calculate markers from cluster IDs
without using the reference labels as identities. This is a recommended
extension, not a step performed by the standalone case-study script:

```{r blinded-label-recovery, purl = FALSE}
pbmc_blinded <- pbmc_final
Idents(pbmc_blinded) <- "seurat_clusters"

cluster_markers <- FindAllMarkers(
  pbmc_blinded,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25,
  verbose = FALSE
)

# Build top-marker lists by the numeric cluster IDs, run CLcompareCluster(),
# lock predictions, and only then attach the reference label table.
cluster_reference <- unique(pbmc_blinded[[]][
  , c("seurat_clusters", "seurat_annotations")
])
```

When several enriched terms are close, do not rank unrelated ontology branches
solely by `CLdepth()`: ancestor count measures specificity only within
ancestor-descendant-comparable terms. Prefer the best-supported term, use depth
to break ties within a lineage, or retain several candidates for review.

## Save the reusable annotation layer and provenance

The standalone script records the declared reference universe and saves the
annotated object and complete R session:

```{r provenance, purl = FALSE}
run_provenance <- data.frame(
  generated_at = format(Sys.time(), tz = "UTC", usetz = TRUE),
  cellontools_version = as.character(packageVersion("CellOnTools")),
  seurat_version = as.character(packageVersion("Seurat")),
  seuratdata_version = as.character(packageVersion("SeuratData")),
  ontology_release = attr(cl_data, "ontology_release"),
  ontology_source = attr(cl_data, "ontology_source"),
  ontology_md5 = unname(tools::md5sum(obo_file)),
  marker_universe = "bundled marker database (universe = NULL)",
  stringsAsFactors = FALSE
)

save_tab(run_provenance, "run_provenance.csv")
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo.txt"))
saveRDS(pbmc_final,
        file.path(out_dir, "pbmc3k_cellontools_annotated.rds"))
```
